---
title: "Comparison of data transformation methods and forecasts with hospitalization data only"
author: "Estee Y Cramer, Serena Wang, Evan Ray, Nick Reich"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Libraries
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(covidHubUtils)
library(fpp3)
library(fable)
```

# Set some global parameters for analysis

```{r}
hosp_data_start_date <- as.Date("2020-08-01")
forecast_date <- as.Date("2021-06-01")
```

# Load and clean truth data 

This code builds "truth" dataset for hospitalizations that goes up through the `forecast_date` specified above (`r forecast_date`)..

```{r, message=FALSE, warning=FALSE}
hosp_truth <- load_truth(
  truth_source = "HealthData",
  target_variable = "inc hosp",
  locations = "25") %>% 
  filter(target_end_date <= forecast_date, target_end_date >= hosp_data_start_date)
```

Turn the truth data above into `tsibbles` and standardize column names.

```{r}
ts_hosp <- hosp_truth %>%
  as_tsibble() %>% 
  select(target_end_date, hosp_val = value) %>% 
  tsibble::fill_gaps()
```

# Define scaled logit
- lower bound = 0 and upper bound = population size in Massachusetts:
```{r}
# define transformation and inverse of transformation
#MA_population <- 6892503
MA_population <- 600
scaled_logit <- function(y, lower_bound = 0, upper_bound = MA_population){
  log((y-lower_bound)/(upper_bound-y))
}

inv_scaled_logit <- function(y,lower_bound = 0, upper_bound = MA_population){
  (upper_bound-lower_bound)*exp(y)/(1+exp(y)) + lower_bound
}

scaled_logit_transformation <- fabletools::new_transformation(scaled_logit, inv_scaled_logit)

```

# Define sqrt transformation
- Note that when using the `sqrt()` transformation for the model below, an error was returned saying "x `upper` can't be lower than `lower`.", which perhaps was arising because of problems taking the square root with numbers between 0 and 1 and the sqrt transformation then making these values larger instead of smaller.
```{r}
# define transformation and inverse of transformation
my_sqrt <- function(x) {sqrt(x)}
truncated_inv_sqrt <- function(x){
  pmax(0,x)^2
}
my_sqrt <- new_transformation(my_sqrt, truncated_inv_sqrt)
```
# Define fourth-root transformation
```{r}
# define transformation and inverse of transformation
fourth_rt<- function(x) {x^(1/4)}
truncated_inv_fourth_rt <- function(x){
  pmax(0,x)^4
}
fourth_rt_transformation <- new_transformation(fourth_rt, truncated_inv_fourth_rt)
```

# Compare variance after data transformation and differencing:
```{r}
ts_hosp %>%
  mutate(scaled_logit_hosp = scaled_logit(hosp_val),
         scaled_logit_hosp_diff = difference(scaled_logit_hosp, 7),
         scaled_logit_hosp_diff2 = difference(scaled_logit_hosp_diff, 1)) %>%
  pivot_longer(-target_end_date) %>%
  ggplot(aes(x=target_end_date, y=value)) +
  geom_line() +
  facet_grid(name~., scales="free_y")

ts_hosp %>%
  mutate(sqrt_hosp = my_sqrt(hosp_val),
         sqrt_hosp_diff = difference(sqrt_hosp, 7),
         sqrt_hosp_diff2 = difference(sqrt_hosp_diff, 1)) %>%
  pivot_longer(-target_end_date) %>%
  ggplot(aes(x=target_end_date, y=value)) +
  geom_line() +
  facet_grid(name~., scales="free_y")

ts_hosp %>%
  mutate(fourth_rt_hosp = fourth_rt_transformation(hosp_val),
         fourth_rt_hosp_diff = difference(fourth_rt_hosp, 7),
         fourth_rt_hosp_diff2 = difference(fourth_rt_hosp_diff, 1)) %>%
  pivot_longer(-target_end_date) %>%
  ggplot(aes(x=target_end_date, y=value)) +
  geom_line() +
  facet_grid(name~., scales="free_y")
```

# Compare ARIMA forecast with hosp data only at different stages 
```{r}
compare_forecast <- function(forecast_date){
  #write model for ARIMA with only hospitalizations 
  scaled_logit_fit <- ts_hosp %>%
    filter(target_end_date >= hosp_data_start_date, 
           target_end_date < forecast_date) %>%
    model(ARIMA(scaled_logit_transformation(hosp_val))) 

  report(scaled_logit_fit)
  
  scaled_logit <- scaled_logit_fit %>% 
      forecast(h=28) %>%
      autoplot(ts_hosp) +
      labs(y = "Hospitalizations", title = paste("ARIMA Forecast Hospitalizations made with scaled logit data prior to ", forecast_date))
  
  #write model for ARIMA with only hospitalizations 
  sqrt_fit <- ts_hosp %>%
    filter(target_end_date >= hosp_data_start_date, 
           target_end_date < forecast_date) %>%
    model(ARIMA(my_sqrt(hosp_val))) 
  
  report(sqrt_fit)

  sqrt <- sqrt_fit %>% 
      forecast(h=28) %>%
      autoplot(ts_hosp) +
      labs(y = "Hospitalizations", title = paste("ARIMA Forecast Hospitalizations made with sqrt data prior to ", forecast_date))
  
   #write model for ARIMA with only hospitalizations 
  fourth_rt_fit <- ts_hosp %>%
    filter(target_end_date >= hosp_data_start_date, 
           target_end_date < forecast_date) %>%
    model(ARIMA(fourth_rt_transformation(hosp_val))) 
  
  report(fourth_rt_fit)

  fourth_rt <- fourth_rt_fit %>% 
      forecast(h=28) %>%
      autoplot(ts_hosp) +
      labs(y = "Hospitalizations", title = paste("ARIMA Forecast Hospitalizations made with fourth root data prior to ", forecast_date))
  
  gridExtra::grid.arrange(scaled_logit, sqrt, fourth_rt, ncol = 1, nrow = 3)
}
```

## heading to the peak
```{r}
compare_forecast(forecast_date = "2020-11-20")
```

## going down from the peak
```{r}
compare_forecast(forecast_date = "2021-01-20")
```

## periods with low hosp value
```{r}
compare_forecast(forecast_date = "2021-05-10")
```