---
title: "VAR modeling for COVID hospitalizations in Massachusetts"
author: "Estee Y Cramer, Serena Wang, Evan Ray, Nick Reich"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Libraries
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(covidHubUtils)
library(fpp3)
library(fable)
```


# Set some global parameters for analysis

```{r}
hosp_data_start_date <- as.Date("2020-08-01")
case_data_start_date <- as.Date("2020-03-01")
forecast_date <- as.Date("2021-06-01")
data_issue_date <- as.Date("2022-02-02")
```

Define fourth root transformation
```{r}
fourth_rt<- function(x) {x^(1/4)}
truncated_inv_fourth_rt <- function(x){
  pmax(0,x)^4
}
fourth_rt_transformation <- new_transformation(fourth_rt, truncated_inv_fourth_rt)
```



# Load and clean truth data 

This code builds two "truth" datasets: one for hospitalizations and one for cases by report date. All three datasets go up through the `forecast_date` specified above (`r forecast_date`). Cases start at `r case_data_start_date` and hospitalizations start at `r hosp_data_start_date`. Filtering out observations with negative values (one observation for cases by report date).

```{r, message=FALSE, warning=FALSE}
hosp_truth <- load_truth(
  truth_source = "HealthData",
  target_variable = "inc hosp",
  locations = "25") %>% 
  filter(target_end_date <= forecast_date, target_end_date >= hosp_data_start_date)

report_date_case_truth <- load_truth(
  truth_source = "JHU",
  target_variable = "inc case",
  temporal_resolution = "daily",
  locations = "25") %>% 
  filter(target_end_date <= forecast_date, target_end_date >= case_data_start_date, value >= 0)
```


Turn the truth data above into `tsibbles` and standardize column names. ALso removing zeroes from case observations

```{r}
ts_report_date_case <- report_date_case_truth %>% 
  select(target_end_date, report_case = value) %>%  #change variable names
  filter(report_case >0) %>%
  mutate(report_case_4throot = fourth_rt_transformation(report_case), 
         report_case_4throot_diff = difference(report_case_4throot, 7)) %>%
  as_tsibble() %>% 
  tsibble::fill_gaps()

ts_hosp <- hosp_truth %>%
  as_tsibble() %>% 
  select(target_end_date, hosps = value) %>%
  mutate(hosps_4throot = fourth_rt_transformation(hosps), 
         hosps_4throot_diff = difference(hosps_4throot, 7)) %>%
  tsibble::fill_gaps()

all_data <- ts_hosp %>%
  left_join(ts_report_date_case) %>%
  filter(!is.na(hosps_4throot_diff) & !is.na(report_case_4throot_diff)) %>%
  fill_gaps() %>%
  mutate(dow = lubridate::wday(target_end_date, label = TRUE),
         isMon = dow=="Mon",
         isTue = dow=="Tue",
         isWed = dow=="Wed",
         isThu = dow=="Thu",
         isFri = dow=="Fri",
         isSat = dow=="Sat")
  
autoplot(all_data, vars(hosps_4throot_diff, report_case_4throot_diff))
```


# Fit hospitalization ARIMA model with no case data inputs

Some exploration of differencing around hospitalization data
```{r}
var_fit <- all_data %>%
  model(fable::VAR(vars(hosps_4throot_diff, report_case_4throot_diff) ~ AR(5) + isMon + isTue + isWed + isThu + isFri + isSat))

var_fit <- all_data %>%
  model(fable::VAR(vars(hosps_4throot_diff, report_case_4throot_diff) ~ AR(1)))


report(var_fit)
```


```{r}
newdata <- tibble(
  target_end_date = seq.Date(forecast_date+1, forecast_date+2, by="1 day"),
  dow = lubridate::wday(target_end_date, label = TRUE),
  isMon = dow=="Mon",
  isTue = dow=="Tue",
  isWed = dow=="Wed",
  isThu = dow=="Thu",
  isFri = dow=="Fri",
  isSat = dow=="Sat"
) %>%
  as_tsibble(index = target_end_date)

tmp <- var_fit %>%
  forecast(h=5) 

autoplot(tmp)
```

